---
title: "lab_2_stat448"
author: "William Hui Chang"
date: "2026-02-26"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Step 1: Load Required Packages

Bla

```{r loading packages}
# install package e1071 (run install.packages just the once if needed),
# or use the Tools -> Install Packages... option from the menu above.
#install.packages("e1071")
#install.packages("ggplot2")
library(e1071)
library(ggplot2)
```



## Step 2: Cars Dataset Description

In industry settings, 
  collaborate with 
    experts 
  clarify 
    each variable’s 
        meaning, 
        significance, and 
        units

use ^ in the R console for details
  ?cars



## Step 3: Setting Up Your Workspace

Check: Use getwd() to view the current directory path.
Set: Use setwd(“P:/Your/Path/Here”) to change it.
Manual Option: Click the “cog” icon under the “Files” tab to set the current directory.



## Step 4: Exploratory Data Analysis (“EDA”)

What is EDA?
  involves analysing datasets to summarise key characteristics, often using visual methods. 
  helping ^ before modelling or hypothesis testing
    uncover patterns, 
      identify anomalies, and 
        validate assumptions 

Origin of EDA
  John Tukey popularised EDA in the 1970s
    through his influential book Exploratory Data Analysis (1977)
    
Key Uses of EDA:
  Data Cleaning: 
    Identify and address missing values, duplicates, or inconsistencies.
  Feature Engineering: 
    Highlight relevant variables and create new features.
  Model Selection: 
    Guide the choice of algorithms for analysis.
  Data Visualisation: 
    Use plots (e.g., histograms, scatterplots, boxplots) to extract insights.
  Error Detection: 
    Spot anomalies that may bias results.
  Hypothesis Generation: 
    Form ideas for confirmatory analysis.



## Step 5: Using EDA to Understand our Data

explore the cars dataset to
  Understand its structure.
  Examine the relationship between speed and distance.
  Assess the distribution of variables.
  Detect any missing values or anomalies.

  Task 1
     investigate the usage of
      dim()
      str()
      head()
      names()


```{r EDA exploring, echo=FALSE}
# use the cars dataset and pass it into a function to understand its dimensions and structure.
# Hint: overwrite the underscore with a function and the dataset
dim(cars)
str(cars)
names(cars)
dim(cars)
```



## Step 6: EDA - Graphing

EDA: Scatter Plots

```{r Scatterplot 1, echo=FALSE}
# Simple Scatterplot
plot(x = cars$speed, y = cars$dist, main = "Dist ~ Speed")
```



```{r Scatterplot 2, echo=FALSE}
# Scatterplot with trend line
scatter.smooth(x = cars$speed, y = cars$dist, main = "Dist ~ Speed")
```


  Task 2
     measure the correlation between speed and stopping distance (dist)



```{r correlation, echo=FALSE}
# Calculate the correlation between speed and dist
cor(cars$speed, cars$dist)
```


## Step 7: EDA: Scatter Plot - Improved Formatting


```{r Scatterplot 3, echo=FALSE}
# An improved Scatter Plot
scatter.smooth(x = cars$speed, y = cars$dist, main = "Stopping Distance vs Speed", xlab = "Vehicle Speed (mph)", ylab = "Stopping Distance (ft)")
```


## Step 8: EDA: Boxplots

par(mfrow()) function allows us to partition our plot window

```{r box plot, echo=FALSE}
# Partition plot window and create boxplots
par(mfrow = c(1, 2))
boxplot(cars$speed, main = "Speed (Mph)", sub = paste("Outlier rows: ", boxplot.stats(cars$speed)$out))
boxplot(cars$dist, main = "Distance (ft)", sub = paste("Outlier rows: ", boxplot.stats(cars$dist)$out))
```


## Step 9: EDA: Density Plots

```{r density 1, echo=FALSE}
# density plot for 'speed'
plot(density(cars$speed),
  main="Density Plot: Speed (Mph)",
  ylab="Frequency", sub=paste("Skewness:", round(e1071::skewness(cars$speed), 2)))
polygon(density(cars$speed), col="red")

```

```{r density 2, echo=FALSE}
# density plot for 'dist'

plot(density(cars$dist),
     main="Density Plot: Distance (ft)",
     ylab= "Frequency", sub=paste("Skewness:", round(e1071::skewness(cars$dist), 2)))
polygon(density(cars$dist), col='blue')

```




