---
title: "lab_2_stat448"
author: "William Hui Chang"
date: "2026-02-26"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Step 0: Create and Understand our Data

Bla

```{r loading packages}
# install package e1071 (run install.packages just the once if needed),
# or use the Tools -> Install Packages... option from the menu above.
#install.packages("e1071")
#install.packages("ggplot2")
# library(e1071)
# library(ggplot2)
```


## Step 1: Create and Understand our Data

Bla

```{r create a dataset}
# create our data using data.frame()
data <- data.frame(x1 = c(0,1,1), y = c(2,2,8))
data
```


Exploratory Data Analysis (EDA)
  Even though our dataset is small, 
    it is still important to 
     graphically visualise our data
  Here we can create a scatter plot 
    to illustrate the relationship between Y and x1.


```{r create a scatterplot}
# create scatter plot
plot(data, xlab = "x1", ylab = "y", main = "Y vs X1", xlim=c(-3, 3), ylim=c(0, 10))
```




## Step 2: Linear Regression Using Linear Algebra

Using linear algebra and partial differentiation
  we minimise the Mean Squared Error (MSE)
    to derive the formula for estimating coefficients

formula
$$
\hat{\beta} = (X^T X)^{-1} X^T Y
$$

To calculate, we need to calculate several Matrices.
We need to:
  (1) Create Design Matrix X
  (2) Transpose our Design Matrix X, so we have XT
  (3) Calculate XTX
  (4) Find the inverse of XTX, so we have (XTX)−1
  (5) Create Y vector,
  (6) Calculate XTY, and finally;
  (7) Calculate (XTX)−1 XTY to find β^


(0) Create: The Intercept Vector

```{r create Intercept Vector}
# create a vector of 1s to use as our intercept
# replicate 1 over the length of our y column from data, i.e, x0 is the same length of y

x0 <- rep(1, length(data$y))
x0
```


(1) Create Design Matrix X

```{r create Design Matrix X}
# Finally, lets combine our x0 and x1 to create our Design Matrix X

X <- as.matrix(cbind(x0, x1 = data$x1))
X

# can check the dimension of a matrix by using 
# dim(X)

```

(2) Create: Response Vector Y


```{r create Response Vector Y}
# create a vector for our Y values

Y <- as.matrix(data$y)
Y

```


(3) Create: Transpose of X


```{r Create Transpose of X}
# We have X and Y. 
# Create X^T
# Note: the transpose of X is found using t()

XT <- t(X)  # X^T
XT
```



## Step 4: Exploratory Data Analysis (“EDA”)

What is EDA?
  involves analysing datasets to summarise key characteristics, often using visual methods. 
  helping ^ before modelling or hypothesis testing
    uncover patterns, 
      identify anomalies, and 
        validate assumptions 

Origin of EDA
  John Tukey popularised EDA in the 1970s
    through his influential book Exploratory Data Analysis (1977)
    
Key Uses of EDA:
  Data Cleaning: 
    Identify and address missing values, duplicates, or inconsistencies.
  Feature Engineering: 
    Highlight relevant variables and create new features.
  Model Selection: 
    Guide the choice of algorithms for analysis.
  Data Visualisation: 
    Use plots (e.g., histograms, scatterplots, boxplots) to extract insights.
  Error Detection: 
    Spot anomalies that may bias results.
  Hypothesis Generation: 
    Form ideas for confirmatory analysis.



## Step 5: Using EDA to Understand our Data

explore the cars dataset to
  Understand its structure.
  Examine the relationship between speed and distance.
  Assess the distribution of variables.
  Detect any missing values or anomalies.

  Task 1
     investigate the usage of
      dim()
      str()
      head()
      names()


```{r EDA exploring, echo=FALSE}
# use the cars dataset and pass it into a function to understand its dimensions and structure.
# Hint: overwrite the underscore with a function and the dataset
dim(cars)
str(cars)
names(cars)
dim(cars)
```



## Step 6: EDA - Graphing

EDA: Scatter Plots

```{r Scatterplot 1, echo=FALSE}
# Simple Scatterplot
plot(x = cars$speed, y = cars$dist, main = "Dist ~ Speed")
```



```{r Scatterplot 2, echo=FALSE}
# Scatterplot with trend line
scatter.smooth(x = cars$speed, y = cars$dist, main = "Dist ~ Speed")
```


  Task 2
     measure the correlation between speed and stopping distance (dist)



```{r correlation, echo=FALSE}
# Calculate the correlation between speed and dist
cor(cars$speed, cars$dist)
```


## Step 7: EDA: Scatter Plot - Improved Formatting


```{r Scatterplot 3, echo=FALSE}
# An improved Scatter Plot
scatter.smooth(x = cars$speed, y = cars$dist, main = "Stopping Distance vs Speed", xlab = "Vehicle Speed (mph)", ylab = "Stopping Distance (ft)")
```


## Step 8: EDA: Boxplots

par(mfrow()) function allows us to partition our plot window

```{r box plot, echo=FALSE}
# Partition plot window and create boxplots
par(mfrow = c(1, 2))
boxplot(cars$speed, main = "Speed (Mph)", sub = paste("Outlier rows: ", boxplot.stats(cars$speed)$out))
boxplot(cars$dist, main = "Distance (ft)", sub = paste("Outlier rows: ", boxplot.stats(cars$dist)$out))
```


## Step 9: EDA: Density Plots

```{r density 1, echo=FALSE}
# density plot for 'speed'
plot(density(cars$speed),
  main="Density Plot: Speed (Mph)",
  ylab="Frequency", sub=paste("Skewness:", round(e1071::skewness(cars$speed), 2)))
polygon(density(cars$speed), col="red")

```

```{r density 2, echo=FALSE}
# density plot for 'dist'

plot(density(cars$dist),
     main="Density Plot: Distance (ft)",
     ylab= "Frequency", sub=paste("Skewness:", round(e1071::skewness(cars$dist), 2)))
polygon(density(cars$dist), col='blue')

```




